Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2025-06-12 Edith Zemanick/Sweat Variability/sweat_test_analysis.qmd
added data from Angela

```{r data cleaning}
#| message: false
# Import CHCO, filter out those with no demo data, rename columns, etc.
chco <- read.csv("./Data_Cleaned/chco_reidentified.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing", "Please reorder testing and re",
    "The received specimen cannot"
  )
)
chco <- chco %>%
  group_by(PatientID) %>%
  filter(Demographics.Available == "TRUE") %>%
  select(
    PatientID, Name, Genotypes1, Genotypes2, Sex, Age.at.Test.in.Years,
    Na, K, Cl, Na_Left, Na_Right, K_Left, K_Right, Cl_Left, Cl_Right
  ) %>%
  rename(Age = Age.at.Test.in.Years) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
# Fix CHCO diagnoses per Angela
chco_diagnoses <- read.csv("./Data_Cleaned/chco_diagnoses.csv")
chco <- full_join(chco, chco_diagnoses, by = join_by(PatientID))
chco$PatientID <- NULL
# For "LESS THAN 10", replace with half the limit of detection
chco[chco == "LESS THAN 10"] <- "9"
chco[, c("Cl", "Cl_Left", "Cl_Right")] <-
  lapply(chco[, c("Cl", "Cl_Left", "Cl_Right")], as.numeric)
# Import NJH and add diagnoses from Katie
njh <- read.csv("./Data_Cleaned/njh_full_list.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing"
  )
)
njh_diagnoses <- read.csv("./Data_Cleaned/njh_diagnoses.csv", na.strings = "")
njh_diagnoses <- njh_diagnoses %>% filter(!is.na(Diagnosis))
njh <- full_join(njh, njh_diagnoses)
# Filter out those with no demo data, rename columns, etc.
njh <- njh %>%
  group_by(Patient.ID) %>%
  filter(Demographics.Available == "TRUE") %>%
  select(
    Patient.ID, Patient.Name, Genotype.1, Genotype.2, Sex, Age.at.test..in.years.,
    Sweat.sodium, Sweat.potassium, Sweat.chloride, Sweat.sodium.left.arm,
    Sweat.sodium.right.arm, Sweat.potassium.left.arm, Sweat.potassium.right.arm,
    Sweat.chloride.left.arm, Sweat.chloride.right.arm, Diagnosis
  ) %>%
  rename(
    Name = Patient.Name, Genotypes1 = Genotype.1,
    Genotypes2 = Genotype.2, Age = Age.at.test..in.years., Na = Sweat.sodium,
    K = Sweat.potassium, Cl = Sweat.chloride, Na_Left = Sweat.sodium.left.arm,
    Na_Right = Sweat.sodium.right.arm, K_Left = Sweat.potassium.left.arm,
    K_Right = Sweat.potassium.right.arm, Cl_Left = Sweat.chloride.left.arm,
    Cl_Right = Sweat.chloride.right.arm
  ) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
njh$Patient.ID <- NULL
# Row bind, fill diagnosis, then remove dupes
df <- rbind(chco, njh)
df <- df %>%
  group_by(Name) %>%
  fill(Diagnosis) %>%
  distinct() %>%
  select(Name, Diagnosis, everything()) %>%
  arrange(Name, Age)
# Use each person's first sweat test
df <- df %>%
  group_by(Name) %>%
  slice_min(Age)
# If there are still duplicates, exclude them
dups <- df$Name[duplicated(df$Name)]
df <- df[!df$Name %in% dups, ]
# For those with left and right, average them
df$Na_mean <- rowMeans(df[, c("Na_Left", "Na_Right")])
df$Na <- coalesce(df$Na, df$Na_mean)
df$Cl_mean <- rowMeans(df[, c("Cl_Left", "Cl_Right")])
df$Cl <- coalesce(df$Cl, df$Cl_mean)
df$K_mean <- rowMeans(df[, c("K_Left", "K_Right")])
df$K <- coalesce(df$K, df$K_mean)
# Get rid of unnecessary columns, remove those with no diagnosis
df <- df %>%
  select(Name, Diagnosis, Genotypes1, Genotypes2, Sex, Age, Na, K, Cl) %>%
  filter(!is.na(Diagnosis))
# Combine CFTR-related disorder and CRMS
df$Diagnosis <- factor(df$Diagnosis,
  levels = c("Cystic Fibrosis", "CFTR-related disorder", "CRMS", "Non-CF"),
  labels = c("Cystic Fibrosis", "CFTR-RD or CRMS", "CFTR-RD or CRMS", "Non-CF")
)
# Calculate Na:Cl ratio
df$`Na:Cl` <- df$Na / df$Cl
# Set table options
mycontrols <- tableby.control(
  numeric.stats = c("N", "Nmiss", "meansd", "medianq1q3", "range")
)
# Save this for Angela et al. to check
write.csv(df,
  file = "./Data_Cleaned/sweat_variability_analysis_dataset.csv",
  row.names = F, na = ""
)
```

>>>>

```{r data cleaning}
#| message: false
# Import CHCO, filter out those with no demo data, rename columns, etc.
chco <- read.csv("./Data_Cleaned/chco_reidentified.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing", "Please reorder testing and re",
    "The received specimen cannot"
  )
)
chco <- chco %>%
  group_by(PatientID) %>%
  filter(Demographics.Available == "TRUE") %>%
  select(
    PatientID, Name, Genotypes1, Genotypes2, Sex, Age.at.Test.in.Years,
    Na, K, Cl, Na_Left, Na_Right, K_Left, K_Right, Cl_Left, Cl_Right
  ) %>%
  rename(Age = Age.at.Test.in.Years) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
# Fix CHCO diagnoses per Angela
chco_diagnoses <- read.csv("./Data_Cleaned/chco_diagnoses.csv")
chco <- full_join(chco, chco_diagnoses, by = join_by(PatientID))
chco$PatientID <- NULL
# Manually add some Na and Cl values that Angela was able to find (using
# numbers so that no PHI is on GitHub)
chco$Na_Right[791] <- 61.3
chco$Na_Right[826] <- 72.4
chco$Na_Left[826] <- 72.4
# For "LESS THAN 10", replace with half the limit of detection
chco[chco == "LESS THAN 10"] <- "9"
chco[, c("Cl", "Cl_Left", "Cl_Right")] <-
  lapply(chco[, c("Cl", "Cl_Left", "Cl_Right")], as.numeric)
# Import NJH and add diagnoses from Katie
njh <- read.csv("./Data_Cleaned/njh_full_list.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing"
  )
)
njh_diagnoses <- read.csv("./Data_Cleaned/njh_diagnoses.csv", na.strings = "")
njh_diagnoses <- njh_diagnoses %>% filter(!is.na(Diagnosis))
njh <- full_join(njh, njh_diagnoses)
# Filter out those with no demo data, rename columns, etc.
njh <- njh %>%
  group_by(Patient.ID) %>%
  filter(Demographics.Available == "TRUE") %>%
  select(
    Patient.ID, Patient.Name, Genotype.1, Genotype.2, Sex, Age.at.test..in.years.,
    Sweat.sodium, Sweat.potassium, Sweat.chloride, Sweat.sodium.left.arm,
    Sweat.sodium.right.arm, Sweat.potassium.left.arm, Sweat.potassium.right.arm,
    Sweat.chloride.left.arm, Sweat.chloride.right.arm, Diagnosis
  ) %>%
  rename(
    Name = Patient.Name, Genotypes1 = Genotype.1,
    Genotypes2 = Genotype.2, Age = Age.at.test..in.years., Na = Sweat.sodium,
    K = Sweat.potassium, Cl = Sweat.chloride, Na_Left = Sweat.sodium.left.arm,
    Na_Right = Sweat.sodium.right.arm, K_Left = Sweat.potassium.left.arm,
    K_Right = Sweat.potassium.right.arm, Cl_Left = Sweat.chloride.left.arm,
    Cl_Right = Sweat.chloride.right.arm
  ) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
njh$Patient.ID <- NULL
# Row bind, fill diagnosis, then remove dupes
df <- rbind(chco, njh)
df <- df %>%
  group_by(Name) %>%
  fill(Diagnosis) %>%
  distinct() %>%
  select(Name, Diagnosis, everything()) %>%
  arrange(Name, Age)
# For those with left and right, average them
df$Na_mean <- rowMeans(df[, c("Na_Left", "Na_Right")], na.rm = T)
df$Na <- coalesce(df$Na, df$Na_mean)
df$Cl_mean <- rowMeans(df[, c("Cl_Left", "Cl_Right")], na.rm = T)
df$Cl <- coalesce(df$Cl, df$Cl_mean)
df$K_mean <- rowMeans(df[, c("K_Left", "K_Right")], na.rm = T)
df$K <- coalesce(df$K, df$K_mean)
# Calculate Na:Cl ratio
df$`Na:Cl` <- df$Na / df$Cl
# Use each person's first sweat test with both sodium and chloride
df <- df %>%
  filter(!is.na(`Na:Cl`)) %>%
  group_by(Name) %>%
  slice_min(Age)
# If there are still duplicates, exclude them
dups <- df$Name[duplicated(df$Name)]
df <- df[!df$Name %in% dups, ]
# Get rid of unnecessary columns, remove those with no diagnosis
df <- df %>%
  select(Name, Diagnosis, Genotypes1, Genotypes2, Sex, Age, Na, K, Cl, `Na:Cl`) %>%
  filter(!is.na(Diagnosis))
# Combine CFTR-related disorder and CRMS
df$Diagnosis <- factor(df$Diagnosis,
  levels = c("Cystic Fibrosis", "CFTR-related disorder", "CRMS", "Non-CF"),
  labels = c("Cystic Fibrosis", "CFTR-RD or CRMS", "CFTR-RD or CRMS", "Non-CF")
)
# Set table options
mycontrols <- tableby.control(
  numeric.stats = c("N", "Nmiss", "meansd", "medianq1q3", "range")
)
# Save this for Angela et al. to check
write.csv(df,
  file = "./Data_Cleaned/sweat_variability_analysis_dataset.csv",
  row.names = F, na = ""
)
```
